---
title: "team.Rmd"
author: "Caleb Purcell"
date: "11/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(readxl)
library(lubridate)
library(dplyr)
library(plotly)
library(ggrepel)
library(maps)
library(mapproj)
```

## Importing Datasets

```{r}
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)

person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

## 1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?

```{r, message = FALSE}
acc$DAY_WEEK = as.factor(acc$DAY_WEEK)
levels(acc$DAY_WEEK) =  c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

acc %>%
  group_by(DAY_WEEK) %>%
  summarise(total_acc = n()) %>%
  arrange(desc(total_acc))
```

    There are 3 days that stand out when looking at this table, and they all happen to be on or near the weekend.  Sunday and Friday have more total accidents compared to the rest of the week, but Saturday has the most with 5873 accidents.

## 2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)

```{r}
fatal_hurt <- person %>%
  filter(INJ_SEV == 4)

head(fatal_hurt)
```

## 3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)

```{r, message = FALSE}
dangerous_cars <- person %>%
  filter(INJ_SEV == 4) %>%
  filter(is.na(MAKE) == FALSE) %>%
  group_by(STATE, MAKE) %>%
  summarise(total_fatal_acc = n()) %>%
  group_by(STATE) %>%
  filter(total_fatal_acc == max(total_fatal_acc))

head(dangerous_cars)
# Vehicle Make Codes: "12" = Ford, "20" = Chevrolet, "37" = Honda, "49" = Toyota 
```

## 4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)

```{r, message = FALSE, warning = FALSE}
# Loading data
glc <- read_xlsx('FRPP_GLC_-_United_StatesSep292020.xlsx')
states <- map_data('state')

# Cleaning data
glc$`State Code` = as.numeric(glc$`State Code`)
dangerous_cars$MAKE = as.factor(dangerous_cars$MAKE)

# Joining Data Frames
text <- dangerous_cars %>%
  left_join(glc %>%
              select(c('State Name', 'State Code')), 
            by = c('STATE' = 'State Code')) %>%
  unique() %>%
  mutate(`State Name` = tolower(`State Name`)) %>%
  left_join(states %>%
              select(long, lat, region),
            by = c('State Name' = 'region'))

# Setting Location & Removing Alaska and Hawaii
location <- text %>% 
  group_by(STATE, MAKE) %>% 
  summarise(long = mean(long, na.rm = TRUE), lat = mean(lat, na.rm = TRUE)) %>%
  filter(!STATE %in% c(2, 15))

# Renaming Levels of MAKE
levels(location$MAKE) =  c("F", "C", "T", "H")

# Visualization
ggplot(states, aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group)) +
  geom_path(aes(group = group), color = 'grey') +
  geom_text(data = location, aes(x = long, y = lat, label = MAKE, color = MAKE)) +
  coord_map() +
  xlab('Longitude') +
  ylab('Latitude') +
  labs(color = 'Company') +
  ggtitle('Most Dangerous Vehicle by State 2017') +
  scale_color_manual(values = c("#FF0000", "#FFFF00", "#00FF00", "#00FFFF"), labels = c("Ford", "Chevrolet", "Toyota", "Honda"))
```

    The most dangerous vehicle is defined by the "dangerous_cars" data frame, which looks at the total fatalities by make for each state. Then the most common fatal make for that state is reported along with the total number of fatalities invloving that make.  From the map we can spot a few trends.  The most fatal make is usually dependent on region.  Ford tends to be more fatal in the South and West, Chevrolet tends to be more fatal across the Midwest, and Toyota/Honda are most fatal in the Northeast.  Overall, Chevrolet appears to be the most dangerous make.
    
## 5. Join the accident and person table (work out which variable(s) to use)

```{r}
acc_person = acc %>%
  select(c(STATE, DAY_WEEK, HOUR)) %>%
  inner_join(person %>% 
               select(c(STATE, SEX, AGE, INJ_SEV)),
             by = 'STATE')

head(acc_person)
```

## 6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.

```{r, message = FALSE}
# Tallying Number of Accidents
acc_person_dow <- acc_person %>%
  group_by(DAY_WEEK, HOUR, SEX) %>%
  summarise(total_acc = n()) %>%
  arrange(desc(total_acc))

# Cleaning Data
acc_person_dow <- acc_person_dow %>%
  filter(SEX %in% c(1, 2),
         HOUR != 99)
acc_person_dow$SEX = as.factor(acc_person_dow$SEX)
levels(acc_person_dow$SEX) =  c("Male", "Female")

# Visualization
acc_person_dow %>%
  ggplot(aes(x = HOUR, weight = total_acc, fill = SEX)) +
  geom_histogram(bins = 24) +
  facet_grid(~DAY_WEEK) +
  ggtitle('Number of Accidents per Hour by Day of the Week') +
  xlab('Hour of Crash') +
  ylab('Number of Accidents') +
  labs(fill = 'Gender')
```

    This visualization shows how the number of accidents increases towards the end of the day for all days of the week.  Accidents are most common during the late evening and early morning hours.

## 7. Now plot a choropleth map of the number of deaths on a county level.

```{r}
# Selecting/Cleaning Data
counties <- map_data("county") %>%
  rename("County Name" = subregion)
county_codes <- glc %>%
  select(c('County Name', 'County Code'))
county_codes$"County Code" <- as.numeric(as.character(county_codes$"County Code"))

# Grouping by County and Joining Data Frames
county_deaths1 <- acc %>%
  group_by(COUNTY) %>%
  summarise('Deaths' = sum(FATALS), .groups = 'drop') %>%
  rename('County Code' = COUNTY)

county_deaths2 <- county_deaths1 %>%
  inner_join(county_codes, by = 'County Code') %>%
  mutate(`County Name` = tolower(`County Name`)) %>%
  distinct() %>%
  inner_join(counties, by ='County Name')

# Visualization
ggplot(counties, aes(x = long, y = lat)) + geom_polygon(aes(group = group)) + geom_polygon(data = county_deaths2, aes(x = long, y = lat, group = group, fill = `Deaths`))+ coord_map() + ggtitle('Auto Accident Deaths by County 2017') + labs(x="Longitude", y="Latitude")
```

    From ths map, we can see there are less auto deaths in rural counties.  This can be seen in Southern Texas as well as parts of the upper Midwest.

## 8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.

```{r}
# Filtering by Season
summer <- acc %>%
  filter(MONTH == 6 | MONTH == 7 | MONTH == 8)
winter <- acc %>%
  filter(MONTH == 1 | MONTH == 2 | MONTH == 12)

# Grouping by State  
summer %>%
  group_by(STATE) %>%
  summarise("Summer Total" = n(), .groups = 'drop')%>%
  head("Summer Total")
winter %>%
  group_by(STATE) %>%
  summarise("Winter Total" = n(), .groups = 'drop')%>%
  head("Winter Total")

# Total Accidents by Season
summer %>%
  tally() %>%
  rename("Summer Total Accidents" = n)
winter %>%
  tally() %>%
  rename("Winter Total Accidents" = n)
```

    According to the total number of accidents, summer is more dangerous than winter (1472 more accidents).  When grouped by state, there are still more accidents in summer months than winter months for most states.  Possible explanations for less winter accidents include less travel and reduced speeds (due to inclement weather).